DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems with PDEs

### 2012-11-29 Paper_May_2012/main.tex
Corrected CGs comments, and fixed pagebreaks

While \deal{} --which at present is the only FE-toolkit to which we provide an 
interface-- leaves the implementation of all high-level algorithms to the user, 
\dope{} is user-focused by delivering
prefabricated tools which requires\todocg{require? Ist plural} adjustments by the user only for parts
connected to his specific problem. The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.

@@

Now we have everything at hand to create an instantiation of the container for the problem data \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{pde} and the \texttt{dof\_handler}.
\begin{lstlisting}
  PDEProblemContainer<LocalPDE> 
             prob_container(pde, dof_handler);                                            
\end{lstlisting}
After that, we add the functional under consideration and, because it is a 
functional acting on the boundary, specify the part of the boundary on which 
the functional operates (this is done via a 'boundary color', where $1$ 
represents $\Gamma_{circ}$).
\begin{lstlisting}
  prob_container.AddFunctional(&functional);
  prob_container.SetBoundaryFunctionalColors(1);
\end{lstlisting}
Next, we incorporate the Dirichlet data. To this end, we instantiate a 
class \texttt{DirichletData} containing the required description of the 
boundary values. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color and a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}
  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0,
                                            comp_mask,
                                            dd);
\end{lstlisting}
Now, we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}
  typedef StatPDEProblem<NEWTON> PDE;
\end{lstlisting}
Here, \texttt{NEWTON} itself is an abbreviation of
\begin{lstlisting}
  typedef NewtonSolver<LINSOLVE> NEWTON;
\end{lstlisting}
The  template parameter \texttt{LINSOLVE} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}
  typedef DirectLinearSolver<MATRIX, VECTOR> LINSOLVE;
\end{lstlisting}
With the types settled, we create the object \texttt{solver}, which needs the discretization (represented here by the \texttt{prob\_container}), the parameter reader as well as the quadrature rules.
\begin{lstlisting}
  PDE pde(prob_container, pr,
                    quad, face_quad);
\end{lstlisting}
Finally, we compute the value $J(u_h)$ by calling:
\begin{lstlisting}
  pde.ComputeReducedFunctionals();
\end{lstlisting}
\todoww{Should we add a complete listing?}
\subsubsection{\texttt{pde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

Let $N$ be the number of degrees of freedom of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e. the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N.
\end{align}
The corresponding Jacobian can be defined analogously.
Now, all we have to supply here are the cell-wise contributions to these terms,
then the \texttt{Integrator} object 
takes care of assembling \eqref{eq:discrete_residual}
and similarly of the matrix for the linearized equation to be solved.

@@

\begin{lstlisting}
  void CellEquation(const CellDataContainer &cdc,
                    VECTOR &local_vector)
  {
    const auto& fe_values = cdc.GetFEValuesState();
    unsigned int dofs_per_cell = cdc.GetNDoFsPerCell();
    unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
To get access to the values and the gradient of the solution in the quadrature 
points we need to extract them from the \texttt{CellDataContainer}:
\todoww{Ich habe jetzt das Arg. in state ge\"andert, das wollten wir sowieso in 
\dope{} umsetzen und ist dann nicht so umst\"andlich}
  \begin{lstlisting}
    cdc.GetValuesState("state", uvalues);
    cdc.GetGradsState("state", ugrads);
 \end{lstlisting}
 In these two lines, we load the values and the gradient of the solution 
during the last Newton iteration in all the quadrature points on the cell 
into the two variables \texttt{uvalues} and \texttt{ugrads}.

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$\todocg{Nochmal: Warum wird $l$ eingefuehrt, wenn unten doch wieder die l2-represenation $f$ verwandt wird?}. Let $I:=[0,T]$ be a time interval with end time point value $T$.
The weak \todocg{formulation?} of such a problem then consists of finding some $u$ such that
\[
\int_I \overline{a}(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a spatial domain $\Omega\in\mathbb{R}^d$. The weak \todocg{formulation?} is then to find 
$u = (v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R})}.
\]
We note that the only difference is the appearance of an additional
time integration, the initial-value, and the term 
\[
\int_I (\partial_t v, \phi)\,dt
\]
so that we can reuse our stationary implementation for all the other 
terms if we rescale them appropriately in the time integration.

@@

\subsection{Goal-Oriented Mesh Refinement for Navier-Stokes}
In this example, we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in Fig.~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn_1,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay is the additional computation of a (linear) dual problem, which is implemented as described in Section~\ref{sec:opt}. In Fig.~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\begin{figure}[hbt]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/ConvergenceNS}}
\caption{Comparison of global and local mesh refinement with respect to the error in the drag-functional $J$.}
\label{fig:NS_comparison}
\end{figure}

@@

The quantities of interest are evaluations of 
$x$- and $y$ displacement at the point $A(0) = (0.6,0.2)$
and the drag and lift forces acting on the cylinder and the elastic beam:
\begin{align}
\label{drag_lift_forces}
(F_D , F_L) 
= {\int_{S_f} \sigma_f \cdot n_f \, ds + 
\int_{\Gamma_i} \sigma_s \cdot n_s \, ds},
\end{align}
where $S_{f}$ denotes the path over the cylinder in the fluid part and
$\Gamma_i$ the interface between the elastic beam and the 
fluid.

Since the problem is fully nonstationary, the 
dynamics are presents in Fig.~\ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Fig.~\ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in \dope{} is correct.

% Bilder von FSI 2 
\begin{figure}[h]
\centering
{\includegraphics[width=5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_mesh8070_scale.png}}
{\includegraphics[width=5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_x_velo8070_scale.png}}
\caption{FSI 2 test case: mesh (left) and velocity profile in vertical 
direction (right) at time $t=16.14s$.}
\label{res:fsi_2_mesh_and_x_velo}
\end{figure}

\begin{figure}
\centering
{\includegraphics[width=5.8cm]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: the deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag $F_D$ and the lift $F_L$ evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}

\newpage
\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam (Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}.
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=1.\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Fig.~\ref{res:mbb}.

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
Finally, the second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.

>>>>

While \deal{} --which at present is the only FE-toolkit to which we provide an 
interface-- leaves the implementation of all high-level algorithms to the user, 
\dope{} is user-focused by delivering
prefabricated tools which require adjustments by the user only for parts
connected to his specific problem. The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.

@@

Now we have everything at hand to create an instantiation of the container for the problem data \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{pde} and the \texttt{dof\_handler}.
\begin{lstlisting}
  PDEProblemContainer<LocalPDE> 
             prob_container(pde, dof_handler);                                            
\end{lstlisting}
After that, we add the functional under consideration and, because it is a 
functional acting on the boundary, specify the part of the boundary on which 
the functional operates (this is done via a 'boundary color', where $1$ 
represents $\Gamma_{circ}$).
\begin{lstlisting}
  prob_container.AddFunctional(&functional);
  prob_container.SetBoundaryFunctionalColors(1);
\end{lstlisting}
Next, we incorporate the Dirichlet data. To this end, we instantiate a 
class \texttt{DirichletData} containing the required description of the 
boundary values. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color and a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}
  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0,
                                            comp_mask,
                                            dd);
\end{lstlisting}
Now, we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}
  typedef StatPDEProblem<NEWTON> PDE;
\end{lstlisting}
Here, \texttt{NEWTON} itself is an abbreviation of
\begin{lstlisting}
  typedef NewtonSolver<LINSOLVE> NEWTON;
\end{lstlisting}
The  template parameter \texttt{LINSOLVE} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}
  typedef DirectLinearSolver<MATRIX, VECTOR> LINSOLVE;
\end{lstlisting}
With the types settled, we create the object \texttt{solver}, which needs the discretization (represented here by the \texttt{prob\_container}), the parameter reader as well as the quadrature rules.
\begin{lstlisting}
  PDE pde(prob_container, pr,
                    quad, face_quad);
\end{lstlisting}
Finally, we compute the value $J(u_h)$ by calling:
\begin{lstlisting}
  pde.ComputeReducedFunctionals();
\end{lstlisting}

\subsubsection{\texttt{pde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

Let $N$ be the number of degrees of freedom (DoFs) of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e., the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N.
\end{align}
The corresponding Jacobian can be defined analogously.
Now, all we have to supply here are the cell-wise contributions to these terms,
then the \texttt{Integrator} object 
takes care of assembling \eqref{eq:discrete_residual}
and similarly of the matrix for the linearized equation to be solved.

@@

\begin{lstlisting}
  void CellEquation(const CellDataContainer &cdc,
                    VECTOR &local_vector)
  {
    const auto& fe_values = cdc.GetFEValuesState();
    unsigned int dofs_per_cell = cdc.GetNDoFsPerCell();
    unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
To get access to the values and the gradient of the solution in the quadrature 
points we need to extract them from the \texttt{CellDataContainer}:
  \begin{lstlisting}
    cdc.GetValuesState("state", uvalues);
    cdc.GetGradsState("state", ugrads);
 \end{lstlisting}
 In these two lines, we load the values and the gradient of the solution 
during the last Newton iteration in all the quadrature points on the cell 
into the two variables \texttt{uvalues} and \texttt{ugrads}.

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side $f$. Let $I:=[0,T]$ be a time interval with end time point value $T$.
The weak formulation of such a problem then consists of finding some $u$ such
 that
\[
\int_I \overline{a}(u)(\phi)\, dt = \int_I (f,\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  
As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a spatial domain $\Omega\in\mathbb{R}^d$. The weak formulation is then to find 
$u = (v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R})}.
\]
We note that the only difference is the appearance of an additional
time integration, the initial-value, and the term 
\[
\int_I (\partial_t v, \phi)\,dt
\]
so that we can reuse our stationary implementation for all the other 
terms if we rescale them appropriately in the time integration.

@@

\subsection{Goal-Oriented Mesh Refinement for Navier-Stokes}
In this example, we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in Fig.~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn_1,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.85\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay is the additional computation of a (linear) dual problem, which is implemented as described in Section~\ref{sec:opt}. In Fig.~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\begin{figure}[hbt]
\centering
\resizebox{0.45\textwidth}{!}{\input{Pictures/ConvergenceNS}}
\caption{Comparison of global and local mesh refinement with respect to the error in the drag-functional $J$.}
\label{fig:NS_comparison}
\end{figure}

@@

The quantities of interest are evaluations of 
$x$- and $y$ displacement at the point $A(0) = (0.6,0.2)$
and the drag and lift forces acting on the cylinder and the elastic beam:
\begin{align}
\label{drag_lift_forces}
(F_D , F_L) 
= {\int_{S_f} \sigma_f \cdot n_f \, ds + 
\int_{\Gamma_i} \sigma_s \cdot n_s \, ds},
\end{align}
where $S_{f}$ denotes the path over the cylinder in the fluid part and
$\Gamma_i$ the interface between the elastic beam and the 
fluid.
% Bilder von FSI 2 
\begin{figure}[h]
\centering
{\includegraphics[width=4.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_mesh8070_scale.png}}
{\includegraphics[width=4.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_x_velo8070_scale.png}}
\caption{FSI 2 test case: mesh (left) and velocity profile in vertical 
direction (right) at time $t=16.14s$.}
\label{res:fsi_2_mesh_and_x_velo}
\end{figure}

Since the problem is fully nonstationary, only a snapshot of the solution
and the mesh are presented in Fig.~\ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Fig.~\ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in \dope{} is correct.

\begin{figure}[H]
\centering
{\includegraphics[width=4.5cm]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: The deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag $F_D$ and the lift $F_L$ evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}

\newpage
\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam (Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}.
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=0.75\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Fig.~\ref{res:mbb}.

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne, initiated by Malte Braack and  Roland Becker \cite{gascoigne}
and RoDoBo, which where initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
Finally, the second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.
